---
title: "Spatial Analysis in R: NICAR 2024"
author: Shreya Vuttaluru 
format: html
editor: visual
---

We'll be using the `sf` package for geospatial functions. This package has many of the same functions available in geospatial software like ArcGIS, QGIS and PostGIS.

Some advantages of using `sf` for spatial analysis:

-   Easy to integrate with other data analysis and cleaning steps
-   Reproduce your scripts
-   Fairly easy visualization
-   Faster processing time for large data sets
-   Free! :money:

### Step 1: Load required packages.

```{r}

### This code will install the packages required for this project if you don't already have them installed.
packages <- c("sf", "tidyverse", "leaflet", "mapview", "janitor", "lubridate", "tidycensus")

for (pkg in packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
  }
}

### Import our packages into our script here.
library(sf)
library(tidyverse)
library(janitor)
library(leaflet)
library(mapview)
library(raster)
library(lubridate)
library(tidycensus)

```

### Step 2: Load in the data sets we'll be working with today. We've pre-cleaned the data so we can focus on the spatial functions. But you'll typically need to clean your data before

-   `shooting_victims.csv` is a data set of unique shooting victims in Baltimore. This means that if there are multiple victims in a shooting, there each row represents a unique victim.

-   `shooting_events.csv` is a data set that contains unique shooting events --- one line for each individual shooting, regardless of how many victims.

-   `high_school_parcels.geojson` is a spatial file of polygons that make up the area of high schools in Baltimore.

Some other types of spatial files you might see: \* `filename.shp` --- An ESRI shapefile. Typically come in a folder with associated files like `.prj, .dbf, and  *`filename.kml`or`filename.kmz`- A Google Earth spatial file *`filename.gdb`— A geodatabase folder of files that manage spatial data. These can be tough to work with, and I recommend handling them in PostGIS or PostgresSQL.  *`filename.rds\` - You can store spatial data in an R data set file

```{r}

### Imporintg shooting victims data
victims <- read_csv("data/shooting_victims.csv") 

### Importing shooting events data
shootings <- read_csv("data/shooting_events.csv") 
  
### Importing a spatial feature of each parcel of each high school in Baltimore. notice we use `st_read()` instead of 'read_csv()'
high_schools <- st_read("data/high_school_parcels.geojson")

```

Let's inspect our shootings data:

```{r}

### First 100 rows
shootings %>%
  head(100)

```

Let's inspect our high schools:

```{r}

### Let's inspect our parcels dataframe
high_schools %>%
  as.data.frame()

```

Let's see what our high schools look like on a map. We will use a function called `mapview()`. It creates an interactive leaflet map that displays our `sf` object.

```{r}

mapview(high_schools)

```

### Step 3: What are some questions we can answer from this data?

-   How many shootings *at* high schools have there been over time?
-   How many shootings *near* high schools have there been over time?
-   Are there particular high schools that experience more shootings than others?
-   What else can you think of?

Let's plan to address the second question --- how many shootings happened *near* high schools in Baltimore?

First, let's take a glance at a subset of where are shootings are. Our data set has over 6,000 rows, which can be tough to visualize all at once. When you have a lot of individual points you'll likely want to filter the data in some way for a meaningful visualization.

Then, we will try to use `mapview()` to view the `csv` we imported on a map. 

```{r}

shootings_subset <- shootings %>%
  head(300)

mapview(shootings_subset)

```

You'll remember that we read in `csv` files of this data, which are not inherently spatial files!

This means we have to convert our data into a spatial object. When we have fields for latitude and longitude, this is a fairly straightforward process.

```{r}

### Transform shootings into a spatial object
spatial_shootings <- shootings %>%
  ### filter out any NA values for longitude and latitude 
  filter(!is.na(longitude) |
         !is.na(latitude)
         ) %>%
  ### use the st_as_sf function to define the spatial aspects of our data set
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326 
  ) 

### Transform victims into a spatial object
spatial_victims <- victims %>%
  ### filter out any NA values for longitude and latitude 
  filter(!is.na(longitude) |
         !is.na(latitude)
         ) %>%
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326 
  )


```

Now we can take a look at where our shootings are using `mapview()` again.

```{r}

shootings_subset <- spatial_shootings %>%
  head(300)

mapview(shootings_subset, 
        #changing the size of the markers
        cex = 3, 
        #changing the color of the markers
        col.regions="red")

```

\*\*\*\*\* A quick side note about Coordinate Reference Systems, also called CRS \*\*\*\*\*\*\*

Coordinate Reference Systems essentially tell our computers what kind of map we want to display our data on. If you're a round-earth truther, you'll know that the earth is *not* flat, and that maps are not perfect representations of what Earth looks like. We set CRS that are the most accurate representations of the areas we are working with.

Truthfully, these are the bane of my existence. [This](https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf) is great resource on which one to use depending on what part of the world you're looking at and the type of analysis you're doing.

If you're in the United States, my favorites are \* 4326 for most visualization-only cases. \* 5070 in any cases where you are calculating perimeter, area, or other mathematical operation. \* You can also cheat and use ChatGPT. 

Datawrapper and Flourish sometimes require 4326.

\*\*\*\*\* END NOTE \*\*\*\*\*\*\*

### Step 4: Buffering

Since we want to analyze how many shootings have taken place near a high school, we have to define what *near* means. Let's pick *3* blocks as our distance.

In Baltimore, we counted 100 meters for each block, plus an additional 50 meters for the immediate street. This is not a block for every city, or the size of every block in Baltimore.

Number of meters to buffer = (Number of blocks \* 100) + 50

*Buffering* involves generating a new polygon (or polygons) that represent an area of a specified distance in a circle around a spatial feature. The code block below creates a three-block radius around each Baltimore high school. We will use the `st_buffer` function and set the size of the buffering by providing a numerical value to an argument called `dist`.

```{r}

buffered_high_schools <- high_schools %>%
  ### let's change the CRS here, since we're doing a mathematical calculation, and working in Meters
  st_transform(5070) %>%
  ### dist is the size of the buffer we are creating
  st_buffer(dist = 350)

```

Now let's look at the output. We can display two different spatial features in one map using `mapview()` if we connect them with a `+`. Here, we will look at the buffered shapes, shaded in blue, and the original high school parcels, shaded in green.

```{r}
  
mapview(buffered_high_schools, col.regions="blue") +
  mapview(high_schools, col.regions="green")

```

Let's also narrow down our shootings data to look at only juveniles.

```{r}

juvenile_spatial_victims <- spatial_victims %>%
  filter(age_range == "juvenile") 

```

### Step 5: Spatial Joins

Spatial Joins are powerful because they allow us to merge geographic data together. Our goal here is to identify which shootings took place inside the buffered ranges we created for each Baltimore high school by using the `st_within()` function. 

```{r}

# Perform a spatial join using st_within
within_buffer <- st_within(juvenile_spatial_victims, buffered_high_schools, sparse = FALSE)

```

Whoops! The CRS don't match. This is a very annoying problem that you will run into frequently. When joining data sets, we always want to make sure that our CRS's are equivalent. When we initially loaded in our `victims` data, we set the CRS to ESPG:4326, but when we buffered our `high_schools`, we changed the CRS to ESPG:5070.

Let's transform our `juvenile_spatial_victims` data to 5070.

```{r}

juvenile_spatial_victims <- juvenile_spatial_victims %>%
  st_transform(crs=5070)

```

Join take two.

```{r}

# Perform a spatial join using st_within. This code creates a matrix 
within_buffer <- st_within(juvenile_spatial_victims, buffered_high_schools, sparse = FALSE)

# We need to apply rowSums to get incidents within any of the school buffers. This will create a data.frame for us. 
shooting_incidents_within_buffer_1 <- juvenile_spatial_victims[rowSums(within_buffer) > 0, ] %>%
  ### count each unique victim only once, even if they are within 
  distinct(cc_number, .keep_all = TRUE)

```

Map this method:

```{r}

map <- mapview(buffered_high_schools, 
               col.region = "blue", # Set the color of the buffered high schools
               alpha.region = 0.3) +
       mapview(high_schools,
               col.region = "green",
               alpha.region = 0.3) +
       mapview(shooting_incidents_within_buffer_1, 
                     col.region = "red", # Set the color of the shooting incidents
                     alpha.region = 0.8,
                     cex = 2, # Set the transparency
                     legend = TRUE) # Enable the legend

# Display the map
map

```

But there's an even easier way to do this! The method above is how we approached it for the story, but using `st_join` and passing through `st_within` as an argument also yields similar results.

```{r}

# Perform a spatial join using st_join with st_within
juvenile_spatial_shootings_within_buffer <- st_join(juvenile_spatial_victims, buffered_high_schools, join = st_within)

# Filter to keep only those incidents that are within a school buffer
# (i.e., where buffered_high_schools data is not NA)
shooting_incidents_within_buffer_2 <- juvenile_spatial_shootings_within_buffer %>%
  filter(!is.na(school_name)) %>%
  distinct(cc_number, .keep_all=TRUE)

```

Map the second method:

```{r}

map <- mapview(buffered_high_schools, 
               col.region = "blue", # Set the color of the buffered high schools
               alpha.region = 0.3) +
       mapview(high_schools,
               col.region = "green",
               alpha.region = 0.3) +
       mapview(shooting_incidents_within_buffer_2, 
                     col.region = "red", # Set the color of the shooting incidents
                     alpha.region = 0.8,
                     cex = 2, # Set the transparency
                     legend = TRUE) # Enable the legend

# Display the map
map

```

Let's plot what these incidents have looked like year over year since 2015.

```{r}

### group and count the incidents by year
shooting_incidents_by_year <- shooting_incidents_within_buffer_2 %>%
  group_by(year) %>%
  summarise(
    count = n()
    )

# Create the ggplot
ggplot(shooting_incidents_by_year, aes(x = year, y = count)) +
  geom_line(color = "#88B04B", size = 1) + # Line plot
  geom_point(color = "#88B04B", size = 2) + # Points on the line
  theme_minimal() + # Minimalist theme
  labs(title = "Number of Shootings Over Time",
       x = "Year",
       y = "Number of Shootings") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) 

```

For extra fun, Census data can also provide some really helpful context and insight into the demographics of the neighborhoods where these shootings occurred.

We found that most of the shootings happened in the Black Butterfly --- an historically red-lined part of Baltimore that has faced decades of disinvestment and disparity.

```{r}

library(tidycensus)
library(tigris)

### load in Baltimore Race Data
black_pop <- get_acs(
  geography = "tract", 
  state = "MD",
  variables = 
    c(total_population = "B03002_001",
      black_pop = "B03002_004"),
  year = 2021,
  geometry = TRUE,
  output = "wide"
  ) %>% 
  mutate(year=("2020")) %>%
  clean_names() %>%
  separate(name, into=c("tract","name","state"), sep=",") %>%
  mutate(
      state = str_trim(state,side="both"),
      name = str_trim(name,side="both")
  ) %>%
  filter(name == "Baltimore city") %>%
  mutate(percent_black = round((black_pop_e/total_population_e) * 100, 2)
    )

black_pop %>%
  as.data.frame()

```

Let's visually inpect the data. We'll display the `percent_black` population, and overlay the shooting incidents.

```{r}

mapview(black_pop,
            zcol = "percent_black",
            col.regions = RColorBrewer::brewer.pal(9, "Greens"),
            alpha.regions = 0.5,
            #at = seq(0, 100),
            legend = TRUE) + 
mapview(shooting_incidents_within_buffer_2,
        col.region = "red",
        cex=2)

```

Now try it on your own! Choose between datasets in Florida or in Baltimore to 

Option 1: Florida Data

* `pinellas_tracts.geojson` is a spatial file of census tracts along with the total population and Black population estimates
* `erp_application.shp` is a shapefile of Environmental Resource Permit applications in Florida since Jan 1, 2023
* `gc_schools_jul23.shp` is a shapefile of schools in Florida. 

Be sure to go into the `option_1` directory for this data. 

```{r}


```

Option 2: Baltimore Data

List datasets here:


Be sure to go into the `option_2` directory for this data. 

```{r}


```

