---
title: "Spatial Analysis in R - 301: NICAR 2026"
author: Ryan Little
format: html
editor: visual

---

This is the third class in our our series on spatial analysis in R. In this class, we will be covering geospatial interpolations. As in our first two classes, we will be using the `sf` package to work with spatial data.

This class assumes you have a basic understanding of spatial data and the `sf` package. If you are new to spatial data in R, let us know and we will try to help catch you up.

## What is interpolation?

Interpolation is the process of estimating values for unknown points within the range of known points. In the context of spatial data, interpolation is used to estimate values for points that are not directly observed.

To do this, we will be working with election data from the last two presidential elections and U.S. Census data. We will use the data from [this map showing the shift toward President Donald J. Trump in the 2024 election](https://www.thebaltimorebanner.com/politics-power/national-politics/maryland-precinct-vote-trump-shift-margin-7GPJDFWDBBHH7OODK2BB77QU3Y/).

Here's the steps we need to take: - Calculate the area of the final geography we want our data in. Today, we will finish in Census tracts. - Calculate the overlaps between the election data and the tracts. - Calculate the area of the overlaps. - Calculate the percentage of the overlaps. - Multiply the election counts by the percentage of the overlap.

## The packages

```{r}

#remove scientific notation
options(scipen=999)

### Run this line to install all packages, if needed. 
#install.packages(c("sf", "tidyverse", "leaflet", "mapview", "janitor", "lubridate", "tidycensus", "lwgeom"))

### Call our packages into our script here.
library(sf)
library(tidyverse)
library(janitor)
library(leaflet)
library(mapview)
library(lubridate)
library(tidycensus)
library(lwgeom)

sf_use_s2(FALSE)

```

## The data

The data we will be using in this class is from the 2024 general election in Baltimore County, Maryland. It includes precinct-level election returns for the presidential election. The data is in GeoJSON format and can be found in the `data` folder in this repository.

```{r}

#read_sf is like read_csv for spatial data (you could also use st_read)
x2024 <- read_sf("data/baltimore-county-election-results-2024.geojson")

#mapview is a simple function to view spatial data on a leaflat map
mapview(x2024)

```

Now let's compare it to data from the 2020 general election. Do these precincts look the same?

```{r}

x2020 <- read_sf("data/baltimore-county-election-results-2020.geojson")

mapview(x2020)

```

There's one more thing we need to tell the story of the 2024 presidential election. We need demographics data from the U.S. Census. Typically, we can get that data using the `tidycensus` package and the shape files from `tigris`. However, some changes in the federal government have made these packages less reliable. We will just import an existing version in the interest of time.

What do you notice about the tracts? How do they compare to the election data?

```{r}

tracts <- read_sf("data/baltimore-county-tracts.geojson")

mapview(tracts)

```

The x2024 election data doesn't include water, but our 2020 precincts and our Census tracts do. One geography that is much bigger than the others will throw off our calculations. We need to clip the data to the same geography.

## Clipping the data

All we need to do is use the `st_intersection` function to clip the data. This will take our existing spatial data and remove any areas that are not in both datasets, sort of like a cookie cutter.

First, we need to make the cookie. If we use our existing dataset of 2024 precincts, we will get intersections inside the tracts, moving ahead to quickly like this.

```{r}

#st_intersection calculates the intersection of two spatial objectsm, in this case creating many new shapes
clipped_tracts <- st_intersection(tracts, x2024)

mapview(clipped_tracts)

```

So we will use `st_union` to combine all the precincts from 2024 into one object. This functions as our "cookie cutter" the shape we will use to match the census tracts and the data that contains water areas.

```{r}

cookie_cutter <- x2024 |>
  #summarise and st_union will combine all the precincts into one object
  summarise(geometry = st_union(geometry))

mapview(cookie_cutter)

```

Now we can use this to cut down the Census tracts and the 2020 precincts.

```{r}

#used this way, st_intersection limits the existing tracts to the area of the cookie cutter
clipped_tracts <- st_intersection(tracts, cookie_cutter)

mapview(clipped_tracts)

```

Now our tracts don't include the water and match the 2024 precincts. Now let's do 2020.

```{r}

clipped_2020 <- st_intersection(x2020, cookie_cutter)

mapview(clipped_2020)

```

## Calculating the area of our final geograpy

Now that we have clipped the data, we can calculate the area of our Census tracts to use as a denominator in our calculations.

```{r}

clipped_tracts <- clipped_tracts |> 
    #st_area calculates the area of a spatial object and returns the value in square meters
    mutate(tract_area = st_area(geometry))

```

## Calculating the overlaps

Now let's calculate the overlaps between the election data and the tracts.

```{r}

#used this way, st_intersection returns new shapes where each shape is a tract that overlaps with a precinct
intersected_2024 <- st_intersection(clipped_tracts, x2024)

mapview(intersected_2024)

```

Notice how our tracts are split into pieces where they overlap with the election data. This is what we want. Now we can calculate the area of the overlaps.

```{r}

intersected_2024 <- intersected_2024 |> 
    mutate(intersect_area = st_area(geometry))

```

Let's take a look at this data.

```{r}

intersected_2024 |>
    dplyr::select(tract_area, intersect_area) |>
    head()

```

All we need to do now is calculate the percentage of the overlaps.

```{r}

intersected_2024 <- intersected_2024 |> 
    #we calculate the overlap percentage by dividing the intersect area by the tract area
    mutate(overlap_pct = intersect_area / tract_area) |>
    #now lets make it a number without a unit of measurment
    mutate(overlap_pct = as.numeric(overlap_pct))

intersected_2024 |>
    dplyr::select(tract_area, intersect_area, overlap_pct) |>
    head()

```

Now we can multiply the election counts by the percentage of the overlap.

```{r}

intersected_2024 <- intersected_2024 |> 
    #we multiply the election counts by the overlap percentage
    mutate(votes_trump_2024 = votes_trump_2024 * overlap_pct,
           votes_dem_2024 = votes_dem_2024 * overlap_pct,
           total_votes_2024= total_votes_2024 * overlap_pct)

intersected_2024 |>
    dplyr::select(tract_area, intersect_area, overlap_pct, votes_trump_2024, votes_dem_2024, total_votes_2024) |>
    head()

```

Now let's get back to whole tracts. All we need to do is group by the tract and sum the votes.

```{r}

grouped_intersected_2024 <- intersected_2024 |> 
    #st_drop_geometry removes the geometry column
    st_drop_geometry() |>
    #we group by the geoid and add up our interpolated votes
    group_by(geoid) |>
    summarise(trump_votes_2024 = round(sum(votes_trump_2024), 0),
              dem_votes_2024 =  round(sum(votes_dem_2024), 0),
              total_votes_2024 = round(sum(total_votes_2024), 0)
              )|>
    #we calculate the share of the vote for Trump
    mutate(trump_share_2024 = trump_votes_2024 / total_votes_2024)

grouped_intersected_2024 |>
  head()

```

And let's catchup 2020 to the same point.

```{r}

grouped_intersected_2020 <- st_intersection(clipped_tracts, x2020) |>
    #calculate intersection area
    mutate(intersect_area = st_area(geometry)) |>
    # calculate the overlapping area
    mutate(overlap_pct = intersect_area / tract_area) |>
    # times the votes by the percentage of the overlap and make it a normal number
    mutate(trump_votes = as.numeric(votes_trump_2020 * overlap_pct),
           biden_votes = as.numeric(votes_dem_2020 * overlap_pct),
           total_votes = as.numeric(total_votes_2020 * overlap_pct)
           )|>
    #removing the geometry column
    st_drop_geometry() |>
    #group by the geoid and add the votes together
    group_by(geoid) |>
    summarise(trump_votes_2020 = round(sum(trump_votes), 0),
              dem_votes_2020 = round(sum(biden_votes), 0),
              total_votes_2020 = round(sum(total_votes), 0)) |>
    #calculate the share of the vote for Trump
    mutate(trump_share_2020 = trump_votes_2020 / total_votes_2020)

```

Now we can join the two data sets together and calculate the shift.

```{r}

trump_shift <- grouped_intersected_2024 |> 
    #join the two datasets by the geoid, a unique identifier for each tract
    left_join(grouped_intersected_2020, by = "geoid") |>
    #calculate the shift by subtracting the 2020 share from the 2024 share, multiplied by 100 to make it a percentage
    mutate(trump_shift = (trump_share_2024 - trump_share_2020)*100)

trump_shift |>
  dplyr::select(geoid, trump_shift, trump_share_2024) |>
  arrange(desc(trump_shift)) |>
  head()


```

The precinct with the biggest shift toward Trump shifted 12% points. Let's find out where it is.

```{r error=TRUE}

#this wont run because trump_shift has no geometry
mapview(trump_shift)

```

The code above doesn't run because it is a regular dataframe. We dropped the geometry when we were making our calculations. We need to join the data back to the tracts sf object that includes the geometry.

```{r}

trump_shift_sf <- clipped_tracts |>
  #join the two datasets by the geoid, a unique identifier for each tract
  left_join(trump_shift, by = "geoid")


#we add the zcol argument to tell mapview what column to use for the color scale
mapview(trump_shift_sf, zcol="trump_shift")

```

And because `trump_shift_sf`is a dataframe, we can filter just the same way we would any map.

```{r}

trump_shift_sf |>
  arrange(desc(trump_shift)) |>
  #this gets the first ten rows
  slice(1:10) |>
  mapview(zcol="trump_shift")

```

## Conclusion

In this class, we learned how to interpolate election data to Census tracts. We used the `sf` package to clip the data and calculate the overlaps between the election data and the tracts. We then calculated the percentage of the overlaps and multiplied the election counts by the percentage of the overlap. Finally, we calculated the shift toward President Donald J. Trump in the 2024 election.

Where in Baltimore County would you send a reporter?
