---
title: "Spatial Analysis in R - 201: NICAR 2026"
author: "Shreya Vuttaluru, updated by Nick McMillan"
date: "2026-03-04"
output: html_document
---

This is the second class in our spatial analysis series. We'll be covering the basics of rasters — image-based data sets that include information extracted from satellites and other aerial tools that are stored in pixels.

This class assumes you have a basic understanding of spatial data and the `sf` package, like shape files and geojsons. If you are new to spatial data in R, let us know and we will try to help catch you up.

We'll be using the terra package (the successor to the raster package) for extracting information from our files and `sf` fore more traditional spatial joins.

::: {.callout-note}

> "Terra is designed to only read into memory those parts of a data set that are necessary for some analysis, raising the possibility of analyzing data sets that are much larger than the availble machine memory"^[<https://pjbartlein.github.io/REarthSysSci/netcdf_terra.html>]

:::

We'll be working with sea surface temperature data used in [this story that tracked how the most intense hurricanes in modern history almost always traverse over record-hot waters.](https://www.tampabay.com/hurricane/2024/11/03/more-hurricanes-are-slamming-gulf-coast-is-this-new-normal/).

We won't be recreating all the findings in this story, since they require scraping and looping through hundreds of files. Instead, we'll walk through an example of how to clean, transform and subset this data in a way that can be applicable to larger projects.


```{r}

### This code will install the packages required for this project if you don't already have them installed. Uncomment this code if you need to install the packages. 

# packages <- c("sf", "tidyverse", "raster", "mapview", "janitor", "lubridate", "tidycensus", "terra", "ncdf4", "stars", "here")
# 
# for (pkg in packages) {
#   if (!require(pkg, character.only = TRUE)) {
#     install.packages(pkg)
#   }
# }

### Import our packages into our script here.
library(sf)
library(tidyverse)
library(janitor)
library(leaflet)
library(mapview)
library(lubridate)
library(tidycensus)
library(here)

### These libraries will help us work with rasters. 
library(ncdf4)
library(terra)
library(stars)

```


## What are rasters, and why should I care about them?

Rasters are digital images that store information in a grid of equally-sized pixels. Environmental agencies and scientists often collect and store data in this format because it allows them to store large summaries of information that is regularly collected about the world in just a single file.

The more pixels we have in a data set, the more detail we can get out of a dataset.

Some examples of popular raster data sets include:
- Land use changes across the United States 
- Sea surface temperature 
- Elevation data


### Okay, but what is really?

Lets create our own raster!

```{r}

ras <- rast(xmin = 0, xmax = 10, ymin = 0, ymax = 10, resolution = 2)

ras

```

What does it look like?

```{r}
plot(ras)
```

We got nothing! Cause there is no value assigned to the cells

So what are we dealing with?

(0,10) --- (10,10)
  |  |  |  |  |
  |  |  |  |  |
  |  |  |  |  |
  |  |  |  |  |
  |  |  |  |  |
(0,0)  --- (10,0)

- The x range is 0 to 10 = 10 degrees wide, divided by resolution of 2 = 5 columns
- The y range is 0 to 10 = 10 degrees tall, divided by resolution of 2 = 5 rows
- We only specified one set of values (no time dimension or multiple variables), so it defaults to 1 layer

How many cells should we have? 
```{r}
ncell(ras)
```

So lets assign some values

```{r}

values(ras) = 1:25

plot(ras)
```

Lets make sure we understand fully whats going:

```{r}
ras
```


- **class**: the type of object — SpatRaster is terra's raster format
- **size**: the grid is 5 rows × 5 columns × 1 layer
- **resolution**: each pixel covers 2 degrees × 2 degrees (x, y)
- **extent**: the bounding box — left, right, bottom, top edges in degrees
- **coord. ref** : the coordinate reference system — WGS84 means standard lat/lon
- **source(s)**: the data lives in RAM, not read from a file on disk
- **name lyr.1**: the name of the layer — auto-assigned since we didn't specify one
- **min value**: the lowest data value stored in any pixel
- **max value**: the highest data value stored in any pixel


## Step 1: the datasets.

Hurricane Milton's path, represented by points every 3 hours as measured by NOAA. 

```{r}

milton_path <- st_read(here("rasters_201/data/milton_path.geojson")) %>%
  st_transform(crs=4326)

glimpse(milton_path)
```

Let's take a quick look at this path. All of these readings by NOAA take place over the storm's entire lifespan, which is several days. 

```{r}

mapview(milton_path, zcol="usa_sshs")

```


Sea surface temperature in 2024.

Raster data can come in a few different file types: some of the most common are NetCDF (.nc) and .TIF files

.TIF files typically store information from a single point in time. In this case, we're looking at ocean temperature data all across the world, as measured only on 10/04/2024. These types of files are typically good for data that doesn't change on a day to day basis — think elevation, or land use. 

Let's take a look at the temperature (measured in celsius) on this October Saturday. 



```{r}

test_tif <- rast(here("rasters_201/data/sea_surface_temp/sst_raster_10042024.tif"))


print(test_tif)

```

::: {.callout-note}

Mapview doesn't handle Terra formats currently. Lets pull out leaflet in the meantime

:::

```{r}

pal <- colorNumeric("inferno", values(test_tif), na.color = "transparent")

leaflet() %>%
  addTiles() %>%
  addRasterImage(test_tif, colors = pal, opacity = 0.8)
```


NetCDF files are like a larger organized collection of data, much like a library with multiple shelves (called "bands") where each shelf holds a specific type of information. Here — each "band" in the NetCDF file represents the temperature on a different day. Think of it like several different .TIF files stacked on top of each other. 

The original version of this file is much larger — it contains information for every single day within 2024. We've subset the file to contain information from just a few days. You can find the original version [here](https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html).


```{r}

nc_sst <- rast(here("rasters_201/data/sea_surface_temp/filtered_sst.nc"))

print(nc_sst)


```

```{r}

# terra parses the time dimension automatically
sst_dates <- as.Date(time(nc_sst))
print(sst_dates)

```


### Step 3: Let's map this object

Can we map this object immediately, like other spatial files? 

```{r error=TRUE}

leaflet() %>%
  addTiles() %>%
  addRasterImage(nc_sst, colors = pal, opacity = 0.8)

```

Since the NetCDF file stacks layers of pixels on top of each other, they're trickier to map right away. So to visualize this object, we're going to need to extract the specific information about the moment in time that we want see. 


Let's extract the data for 10-04-2024. Feel free to try a couple of dates in `target_date`, and see what happens. Why do some dates work, but not others? 

```{r}

target_date <- as_date("2024-10-04")
date_index <- which(sst_dates == target_date)

# subset to just that layer
projected_sst <- nc_sst[[date_index]]

```


```{r}

pal <- colorNumeric("inferno", values(projected_sst), na.color = "transparent")

leaflet() %>%
  addTiles() %>%
  addRasterImage(projected_sst, colors = pal, opacity = 0.8)

```

Let's transform the data for a single day, so we can view it. 

```{r}

# Extract SST data for that specific day
sst_data <- ncvar_get(nc_sst, "sst", 
                       start = c(1, 1, date_index), 
                       count = c(-1, -1, 1))

# Get longitude and latitude values
lon <- ncvar_get(nc_sst, "lon")
lat <- ncvar_get(nc_sst, "lat")

# Convert SST data into a raster object
sst_raster <- raster::raster(t(sst_data), 
                              xmn = min(lon), xmx = max(lon), 
                              ymn = min(lat), ymx = max(lat))

```

Before we project this into an object we can map, let's check how big the pixels are:

Note: this size is in DEGREES, not meters. We're not calculating area today, but in cases where you are calculating area from pixel files, be sure to convert your calculations to traditional measures of area. 

```{r}

res(sst_raster)

```

Now, let's project it into WGS84

```{r}

# Set the correct CRS to WGS84 (EPSG:4326) — the same CRS as we set milton path to earlier
crs(sst_raster) <- CRS("+proj=longlat +datum=WGS84 +no_defs")

mapview(sst_raster)

```

That looks wrong! Let's rotate and project this right-side up: 

The final map should show a logical basemap, and map sea surface temperature in degrees celsius, much like the TIF file we mapped first. 

```{r}

# Rotate and flip the raster to ensure the orientation is correct
projected_sst <- rotate(sst_raster) |>
  flip(direction = 'y')

# View the rotated raster using mapview
mapview(projected_sst)

```


Now we have sea surface temperature data at every point in the ocean on October 4th, 2024. 

We also know that then Tropical Depression Milton had formed in the Gulf of Mexico on this day, so we can track both the storm's movement and the temperature of the water. 

```{r}

milton_points_oct_4 <- milton_path |>
  mutate(
    date = as.Date(iso_time)
  ) |>
  filter(date == "2024-10-04")

mapview(milton_points_oct_4, zoom=4, zcol="usa_sshs")

```

How hot was the Gulf of Mexico on October 4th, while NOAA was measuring the wind speed? 

We'll use the `extract()` function within the raster package, which takes in points and then extracts the information from the corresponding raster. 

```{r}

### extract the coordinates on october 4th
milton_points <- st_coordinates(milton_points_oct_4$geometry)

### extract the SST 
sst_values_oct_4 <- raster::extract(projected_sst, milton_points)

milton_points_oct_4$sst <- sst_values_oct_4

milton_points_oct_4 %>%
  dplyr::select(iso_time, usa_sshs, usa_wind, sst) %>%
  st_drop_geometry()

```

In our overall `milton_path` and `nc_sst` dataset, we have 7 days of sea surface temperature data, and 7 days in which Milton was tracked by NOAA. These should be the same dates, and we can check by printing a list: 

```{r}

milton_path_dates <- milton_path |>
  mutate(
    date = as.Date(iso_time)
  ) |>
  group_by(date) |>
  summarize(
    count = n()
  ) %>%
  pull(unique(date))

print(milton_path_dates)
print(sst_dates)

```

What if we want to know how hot the water was during all the points that NOAA did a readings, not just during a single day? To do this, let's set up a loop! 

First, let's re-extract the correct dates from the netcdf — remember that the initial units were "days since 1980-01-01". 
Fun fact: that's because the modern satellite era begins in 1980! 

```{r}

# get the time indices and dates from nc_sst (we did this step earlier, too)
time_var <- ncvar_get(nc_sst, "time")
time_units <- ncatt_get(nc_sst, "time", "units")
ref_time <- as.Date(substr(time_units$value, 12, 21))
sst_dates <- ref_time + time_var

# Get lon/lat once outside the loop
lon <- ncvar_get(nc_sst, "lon")
lat <- ncvar_get(nc_sst, "lat")

# esure milton_path has date column that we can iterate through
milton_path <- milton_path %>%
  mutate(date = as.Date(iso_time))

```

For every date in `milton_path`, we'll get the corresponding SST layer within the netcdf file. Then, we'll convert it into a raster object, project it appropriately, and extract the sea surface temperature at the point that was tracked in the storm's path. 

```{r}

# initialize a list to store results
sst_results <- list()

# Loop through each unique date in milton_path
for (target_date in unique(milton_path$date)) {
  # Find the matching index in the netCDF file's time dimension
  date_index <- which(sst_dates == target_date)
  
  if (length(date_index) == 0) {
    warning(paste("No SST data found for date:", target_date))
    next
  }
  
  # extract SST data for this date
  sst_data <- ncvar_get(nc_sst, "sst", 
                        start = c(1, 1, date_index), 
                        count = c(-1, -1, 1))
  
  # convert SST data into a raster
  sst_raster <- raster(t(sst_data), 
                      xmn = min(lon), xmx = max(lon), 
                      ymn = min(lat), ymx = max(lat))
  
  # set CRS and correct orientation
  crs(sst_raster) <- CRS("+proj=longlat +datum=WGS84 +no_defs")
  projected_sst <- rotate(sst_raster) %>%
    flip(direction = 'y')
  
  # get storm points for this date
  date_storms <- milton_path %>% 
    filter(date == target_date)
  
  # extract SST values at storm points
  sst_values <- raster::extract(projected_sst, date_storms)
  
  # store extracted SST in the data
  date_storms$sst <- sst_values
  sst_results[[length(sst_results) + 1]] <- date_storms
}

# combine results into a single dataframe
final_results <- bind_rows(sst_results)

# view results
final_results %>%
  arrange(date) %>%
  dplyr::select(date, iso_time, usa_sshs, usa_wind, sst) %>%
  st_drop_geometry() %>%
  head(10)

```

Finally, let's map the sea surface tempature along the path! 

```{r}

mapview(final_results, cex = "usa_sshs", zcol = "sst")

```
